---
title: "Assessing Premier League Player Valuation: The Impact of Age, Position, Playtime and Other Variables on Market Value" 
author: "Stat Warriors"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
library(dplyr)
library(corrplot)
library(Hmisc)
library(car)
library(olsrr)
library(caret)
library(tidyverse)

soccerdata<-read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTTPKPRxczOS1SkmereE95_6ujc9b6oEcjJGCOSsPQAD-Cwc82uoC-c1rWLbYwqX2s_MftoTDTldHrf/pub?gid=443710304&single=true&output=csv",header=T)
```


\begin{center}

\includegraphics{/Users/sakethvangara/Downloads/soccer_ball_holding_money.jpeg}


\end{center}

\newpage

## Introduction

## Methods and Analysis

## Results

## Conclusions

\newpage

## Appendix A: Data Dictionary

|    Variable Name     | Abbreviated Name |              Description              |
|:------------------:|:----------------:|:--------------------------------:|
| Variable 1 Full Name |   Abbreviated1   | Write the description of the variable |
| Variable 2 Full Name |   Abbreviated2   | Write the description of the variable |

\newpage

## Appendix B: Data Rows

```{r, echo=FALSE}
# Use head() function to output first several rows
head(soccerdata)
```

\newpage

## Appendix C: Tables and Figures

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries 
## that will not be included in your appendix
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from EDA, assumptions, and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

#reduced data table (removed variables not being used)
soccerdatared <- select(soccerdata, Valuation, Age, Position, Club.Ranking.Group, minutes_played_overall, England, appearances_overall, goals_overall, assists_overall, clean_sheets_overall, conceded_overall, goals_involved_per_90_overall, min_per_match)

#data table with only quantitative variables
soccerdataquant <- select(soccerdata, Valuation, Age, minutes_played_overall, appearances_overall, goals_overall, assists_overall, clean_sheets_overall, conceded_overall, goals_involved_per_90_overall, min_per_match)

#RESPONSE TRANSFORMATION STUFF
#log response transformation
soccerdatared$logval<-log(soccerdatared$Valuation)
hist(soccerdatared$logval, xlab="Valuation", main="Histogram of log Valuation") 
soccerdataquant$logval<-log(soccerdataquant$Valuation)

#EDA ON LOG VERSION OF RESPONSE
plot(soccerdatared$Age, soccerdataquant$logval,xlab="Age",ylab="log of Valuation", main=)
plot(soccerdatared$minutes_played_overall, soccerdataquant$logval, xlab="Minutes Played Overall", ylab="log of Valuation")

#Box plots for Qualitative
boxplot(soccerdatared$logval~Position,soccerdata, ylab="log of Valuation")
boxplot(soccerdatared$logval~Club.Ranking.Group,soccerdata, ylab="log of Valuation")
boxplot(soccerdatared$logval~England,soccerdata, ylab="log of Valuation")

#updated model with significant quantitative and qualitative predictors
soccermod4<-lm(logval~clean_sheets_overall + Age + goals_involved_per_90_overall + min_per_match + Club.Ranking.Group,data=soccerdatared)

# REGRESSION ASSUMPTIONS
#Residuals Plots of explanatory variables vs residuals
residualPlots(soccermod4,tests=F)

#Residual vs Fitted and QQ plot
plot(soccermod4, which=c(1,2)) 

#histogram of residuals
hist(residuals(soccermod4))

# INFLUENTIAL DIAGNOSTICS

#Cooks Distance Thresholds
plot(soccermod4,which=4) 

#Leverage vs Studentized Residuals
influencePlot(soccermod4,fill=F)

#Deleted Studentized Residuals vs PRedicted values
ols_plot_resid_stud_fit(soccermod4)


#We can use various functions to store and view these statistics for all observations
rstudent(soccermod4) #rstudent gives the deleted studentized residuals
hatvalues(soccermod4)
cooks.distance(soccermod4)
dffits(soccermod4)
dfbetas(soccermod4)

# FINAL MODEL
subsetsoccerdatared<-soccerdatared[-c(24,25,38,82,91,97),] #the -c means remove those specific observations
finalsoccermod<-lm(logval~clean_sheets_overall + Age + goals_involved_per_90_overall + min_per_match + Club.Ranking.Group,data=subsetsoccerdatared)
summary(finalsoccermod)

# CROSS VALIDATION

# Check for missing values
colSums(is.na(soccerdatared))

# Remove rows with missing values
soccerdatared <- na.omit(soccerdatared)

# Confirm no missing values remain
colSums(is.na(soccerdatared))


set.seed(125)
train.kfold <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

### Training
model.kfold <- train(logval~clean_sheets_overall + Age + goals_involved_per_90_overall + min_per_match + Club.Ranking.Group, data = soccerdatared, method = "lm", trControl = train.kfold)

# Present results
model.kfold

```

\newpage

## Appendix D: References

### Background

1. DiBlasi, L. (2022, April). Footballer Valuations: Valuing World-Class Football Players Against Transfer Fees [Review of Footballer Valuations: Valuing World-Class Football Players Against Transfer Fees]. Bryant University. https://digitalcommons.bryant.edu/cgi/viewcontent.cgi?article=1042&context=honors_economics

2. Metelski, A. (2021, April 30). Factors affecting the value of football players in the transfer market. ResearchGate. https://www.researchgate.net/publication/351335017_Factors_affecting_the_value_of_football_players_in_the_transfer_market

3. Poli, R., Ravenel, L., & Besson, R. (2020, March). Scientific evaluation of the transfer value of football players. Monthly Report 53. https://football-observatory.com/IMG/sites/mr/mr53/en/

4. Zhou, W. (2019). Which Factors Decide the Market Value of Soccer Players: An Empirical Evidence from European League. Clausius Press. https://www.clausiuspress.com/conferences/LNEMSS/EMSD%202019/19EMSD057.pdf

### Data

1.  Players CSV / Football Stats Database to CSV | FootyStats. (2019). Footystats.org. https://footystats.org/download-stats-csv

2. Premier League 18/19. (2018). Transfermarkt.us. https://www.transfermarkt.us/premier-league/startseite/wettbewerb/GB1/saison_id/2018

### Supplemental Code and Analysis Help

1.  Cheng, M. (2023, February 15). Simple examples of cross-validation. RPubs. https://rpubs.com/muxicheng/1004550 
